Verifying Android Applications using Java PathFinder

### 2012-09-04 Papers/paper.tex
Updating paper changes

\author{%
\alignauthor Heila van der Merwe, Brink van der Merwe and Willem Visser\\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{University of Stellenbosch}\\
       \affaddr{Private Bag X1 Matieland}\\
       \affaddr{South Africa, 7602}
       \email{hvdmerwe@cs.sun.ac.za, abvdm@cs.sun.ac.za, wvisser@cs.sun.ac.za }
}

@@

\keywords{Mobile Application, Java PathFinder, JPF, Android, Verification, Testing}

@@

Although software testing is so important, it is often neglected due to its complex and time consuming process. Android
applications face additional challenges when it comes to application testing. Firstly, they have an event based design which means that
their application flow is driven by graphical user interface (GUI) and system events. Secondly, Android applications are developed in a
custom implementation of the Java Application Programming Interface (API)  adopted from the Apache Harmony~\cite{harmony} project. The
compiled applications can only be executed on a special virtual machine (VM) called the Dalvik VM that runs on Android
devices~\cite{dalvik}.

@@

Although Android applications and Java desktop applications are designed for completely different Java VMs, they are both built on an
implementation of the Java API. As a consequence, Android applications contain many of the same errors as Java applications. These defects
include, but are not limited to, concurrency issues and common runtime exceptions such as NullPointer Exceptions. It follows that existing
Java testing frameworks can be adapted to verify Android applications.

% what are we going to do and why 
This paper describes an extension to Java Pathfinder (JPF)~\cite{JPFDocs} that enables the automatic verification of Android applications on
the standard Java JVM.

The next section will provide an overview of JPF and Android and how JPF can be used to test Android applications. Thereafter
the development of the JPF-ANDROID tool will be discussed concluding with two case studies.

\section{Design}
\subsection{Java PathFinder}
JPF is an automated, open source, analysis engine for Java applications~\cite{JPFDocs}. It is implemented as an explicit state model checker
that includes mechanisms to model Java classes and native method calls (Model Java Interface - MJI Environment), track byte code execution
and listen for property violations. Additionally, JPF's design encourages developers to create extensions to the framework. Currently there
exist many extensions including a symbolic execution extension (JPF-SYMBC), data race
detector (JPF-RACEFINDER) and an abstract window toolkit (AWT) extension (JPF-AWT)~\cite{JPF-AWT}.

JPF-AWT allows the model checking of AWT applications. AWT applications are event driven and based on a single
threaded, message queue design. All application events are put in a message queue and then handled by the main thread of the application,
called the EventDispatchThread. JPF-AWT introduced the idea of using a simple event script file to write sequences of user inputs to drive
the application execution. JPF-AWT models the EventDispatchThread so that when the message queue is empty, it requests an event from the
script file simulating the event occurring.

JPF-ANDROID makes use of JPF's extension mechanisms to model the Android application framework so that Android applications can run on the
JVM. It then extends JPF-AWT's input model to simulate user and system input to drive the application flow. One of the advantages of
extending JPF is that it has been rigorously tested and can successfully detect many common defects in Java applications. As soon as Android
applications can run on JPF-ANDROID, common software errors are automatically detected.

@@

As shown in figure \ref{fig:android}, the Android OS is built on top of a modified Linux kernel. The kernel provides a layer of abstraction
on top of its low level functionality such as process, memory, user, network and thread management. On top of the Android kernel is a set of
native C-libraries. The Dalvik VM is a custom, optimised version of the JVM. For security reasons each Android application is run as a
separate process in its own Dalvik VM instance~\cite{AndroidSecurity}.  Hence, applications can only communicate with each other and the
with the Android framework using Android's Binder inter-process communication (IPC) mechanism~\cite{Binder}.

@@

Android has one main application called the system process. The system process contains services responsible for performing the main
tasks of the system including:
\vspace{-10pt}
\begin{description}
 \item [ActivityManager] manages the life-cycle and interaction of all the activities running on the system
 \item [WindowManager] allows applications to draw on the screen and forwards UI input to the application.
 \item [PackageManager] stores information on the application packages installed on the device
\end{description}
\vspace{-5pt}
All Android applications follow a single-threaded design in which the main thread of the application handles all application
events~\cite{AndroidDocs}. This structure is commonly used by many UI frameworks since it becomes too complex to make all UI classes thread
safe~\cite{SingleThread}. In Android, this main thread is called the \texttt{Looper} thread. The \texttt{Looper} has a message queue
attached to it containing all application events to be dispatched. UI and system events are scheduled on the main \texttt{Looper} by adding
them to the message queue. The \texttt{Looper} is responsible for continuously looping through these messages and handling them
appropriately. This could include updating a widget, loading a new activity or processing an Intent.

@@

\section{Development}
\subsection{JPF-ANDROID architecture}
% \begin{figure}
% \centering
% \epsfig{file=arch.eps, width=3.5in,}
% \caption{JPF-ANDROID architecture}
% \label{fig:arch}
% \end{figure}
%JPF-ANDROID BASED ON JPF_AWT  why and how does it work?
As discussed above, both Android and AWT applications have a single-threaded application design. That is why JPF-ANDROID's architecture
is based on JPF-AWT. JPF-ANDROID models Android's message queue by using JPF's MJI
environment. When the \texttt{Looper} thread requests a new message from the message queue and it is empty, a call to the native
\texttt{JPF\_MessageQueue} class requests the next event from the scripting environment. The \texttt{JPF\_MessageQueue} class classifies
events as either UI or system events. UI events are directed to and handled by the native \texttt{JPF\_Window} class and system events by
the native \texttt{JPF\_ActivityManager} class.

@@

%how and why was this scripting environment adapted
Android applications contain multiple windows - one for each Activity. This complicates the script input as each window has its own unique
set of view components, hence, a unique input sequence. Furthermore, the application flow can switch between these Activities
at any point of execution. In other words, if we do not know in what Activity the application is, we can not script the following events.

>>>>

\author{%
\alignauthor Heila van der Merwe, Brink van der Merwe and Willem Visser\\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{University of Stellenbosch}\\
       \affaddr{Private Bag X1 Matieland}\\
       \affaddr{South Africa, 7602}
       \email{\{hvdmerwe, abvdm, wvisser\}@cs.sun.ac.za }
}

@@

\keywords{Mobile Application, Java Pathfinder, JPF, Android, Automatic Verification, Testing, Model Checking}

@@

Although software testing is so important, it is often neglected due to its complex and time consuming process. Android
applications face additional challenges when it comes to application testing. Firstly, they have an event based design which means that
their application flow is driven by graphical user interface (GUI) and system events~\cite{Hu:2011}. Secondly, Android applications are
developed in a
custom implementation of the Java Application Programming Interface (API)  adopted from the Apache Harmony~\cite{harmony} project. The
compiled applications can only be executed on a special virtual machine (VM) called the Dalvik VM that runs on Android
devices~\cite{dalvik}.

@@

Although Android applications and Java desktop applications are designed for completely different Java VMs, they are both built on an
implementation of the Java API. As a consequence, Android applications contain many of the same errors as Java applications. These defects
include, but are not limited to, concurrency issues and common run-time exceptions such as null pointer exceptions. It follows that
existing Java testing frameworks can be adapted to verify Android applications.

% what are we going to do and why 
This paper describes an extension to Java Pathfinder (JPF)~\cite{JPFDocs} that enables the automatic verification of Android applications on
the standard Java JVM. The next section will provide an overview of JPF and Android and how JPF can be used to test Android applications.
Thereafter, the development of the JPF-ANDROID tool will be discussed concluding with two case studies.

\section{Design}
\subsection{Java Pathfinder}
JPF is an automated, open source, analysis engine for Java applications~\cite{JPFDocs}. It is implemented as an explicit state model checker
that includes mechanisms to model Java classes and native method calls (Model Java Interface - MJI Environment), track byte code execution
and listen for property violations. Additionally, JPF's design encourages developers to create extensions to the framework. Currently there
exist many extensions including a symbolic execution extension (JPF-SYMBC), data race
detector (JPF-RACEFINDER) and an abstract window toolkit (AWT) extension (JPF-AWT)~\cite{JPF-AWT}.

JPF-AWT allows the model checking of AWT applications. AWT applications are event driven and based on a single
threaded, message queue design. All application events are put in a message queue and then handled by the main thread of the application,
called the \texttt{EventDispatchThread}. JPF-AWT introduced the idea of using a simple event script file to write sequences of user inputs
to drive the application execution. JPF-AWT models the \texttt{EventDispatchThread} so that when the message queue is empty, it requests an
event from the script file simulating the event occurring.

JPF-ANDROID makes use of JPF's extension mechanisms to model the Android application framework so that Android applications can run on the
JVM. It then extends JPF-AWT's input model to simulate user and system input to drive the application flow. One of the advantages of
extending JPF is that it has been rigorously tested and can successfully detect many common defects in Java applications. When we run
Android applications on JPF-ANDROID, common software errors are automatically detected.

@@

As shown in figure \ref{fig:android}, the Android OS is built on top of a modified Linux kernel. The kernel provides a layer of abstraction
on top of its low level functionality such as process, memory, user, network and thread management. On top of the Android kernel is a set of
native C-libraries. The Dalvik VM is a custom, optimised version of the JVM. For security reasons each Android application is run as a
separate process in its own Dalvik VM instance~\cite{AndroidSecurity}.  Hence, applications can only communicate with each other and with
the Android framework using Android's Binder inter-process communication (IPC) mechanism~\cite{Binder}.

@@

Android has one main application called the system process. The system process contains services responsible for performing the main
tasks of the system including:
\vspace{-10pt}
\begin{description}
 \item [ActivityManager] manages the life-cycle and interaction of all the activities running on the system
 \item [WindowManager] allows applications to draw on the screen and forwards UI input to the application.
 \item [PackageManager] stores information on the application packages installed on the device
\end{description}
\vspace{-5pt}
All Android applications follow a single-threaded design in which the main thread of the application handles all application
events~\cite{AndroidDocs}. This structure is commonly used by many UI frameworks since it becomes too complex to make all UI classes thread
safe~\cite{SingleThread}. In Android, this main thread is called the \texttt{Looper} thread. The \texttt{Looper} has a message queue
attached to it containing all application events to be dispatched. UI and system events are scheduled on the main \texttt{Looper} by adding
them to the message queue. The \texttt{Looper} is responsible for continuously looping through these messages and handling them
appropriately. This could include updating a widget, loading a new activity or processing an \texttt{Intent}.

@@

\section{Development}
\subsection{JPF-ANDROID architecture}
% \begin{figure}
% \centering
% \epsfig{file=arch.eps, width=3.5in,}
% \caption{JPF-ANDROID architecture}
% \label{fig:arch}
% \end{figure}
%JPF-ANDROID BASED ON JPF_AWT  why and how does it work?
As discussed above, both Android and AWT applications have a single-threaded application design. That is why JPF-ANDROID's architecture
is based on JPF-AWT.
JPF- ANDROID models Android's message queue by using JPF's MJI
environment. When the \texttt{Looper} thread requests a new message from the message queue and it is empty, a call to the native
\texttt{JPF\_MessageQueue} class requests the next event from the scripting environment. The \texttt{JPF\_MessageQueue} class classifies
events as either UI or system events. UI events are directed to and handled by the native \texttt{JPF\_Window} class and system events by
the native \texttt{JPF\_ActivityManager} class.

@@

%how and why was this scripting environment adapted
Android applications contain multiple windows - one for each Activity. This complicates the script input as each window has its own unique
set of view components, hence, a unique input sequence. Furthermore, the application flow can switch between these Activities
at any point of execution. In other words, if we do not know which Activity has the focus (or is in it's resumed state), we can not script
the following events.
